---
title: "Task 07 - R Notebook"
output:
  html_document:
    keep_md: TRUE
  html_notebook: default
---

##Introduction

This R Markdown Notebook has been created for as part of Task 06 for the course GEOL 590. This task is an introduction to using the **dplyr** package with the built-in datasets **nycflights13** and **babynames**.

##Assigned tasks

* Using the `nycflights13::weather`:
    * Determine whether there are any clear outliers in wind speed (`wind_speed`) that should be rejected. If so, filter those bad point(s) and proceed.
    * What direction has the highest median speed at each airport? Make a table and a plot of median wind speed by direction, for each airport. *Optional fun challenge: If you like, this is a rare opportunity to make use of `coord_polar()`.*
    
```{r}
library(tidyverse) #Loads the packages associated with the tidyverse
library(nycflights13) #Loads the nycflights dataset

#Calculates median wind speed of data by airport and direction for use in table and plot
median_wind_speed <- weather %>%
  group_by(origin, wind_dir) %>%
  select(origin, wind_dir, wind_speed) %>%
  #Filters out large outlier in wind speed as well as data for which there is no wind direction specified
  filter(wind_speed < 50, wind_dir != "NA", wind_dir != "0") %>%
  summarise(Median.Wind.Speed = median(wind_speed, na.rm = TRUE)) %>%
  rename(Wind.Direction = wind_dir, Origin = origin) %>%
  print()


library(DT) #Loads the DT package which provides an R interface for the JavaScript library DataTables

#Creates a datatable of median wind speed data using the DT package
#This interactive datatable is only able to be seen when opening as an HTML file
datatable(median_wind_speed, class = "cell-border stripe", colnames = c("Airport", "Wind Direction", "Median Wind Speed (mph)"))

#Makes a plots of median wind speed as a function of wind direction by airport
wind_speed_plot <- ggplot(median_wind_speed, aes(Wind.Direction, Median.Wind.Speed, color = Median.Wind.Speed)) +
  geom_point() +
  facet_wrap(~Origin) +
  labs(x = "Wind direction", y = "Median wind speed (mph)")

print(wind_speed_plot)

```

* Using `nycflights13::flights` and `nycflights13::airlines`:
    * Make a table with two columns: airline name (not carrier code) and median distance flown from JFK airport. The table should be arranged in order of decreasing mean flight distance. Hint: use a `_join` function to join `flights` and `airlines`.
    
```{r}

#Joins the flights and airlines datatables together so full airline names may be displayed
flights_and_airlines <- inner_join(flights, airlines, by = "carrier")

#Calculates the median distance of flights that leave from JFK by airline
median_dist_by_airline <- flights_and_airlines %>%
  select(name, origin, distance) %>%
  filter(origin == "JFK") %>%
  group_by(name) %>%
  summarise(Median.Distance = median(distance)) %>%
  arrange(Median.Distance) %>%
  print()

```

* Make a *wide-format* data frame that displays the number of flights that leave Newark ("EWR") airport each month, from each airline

```{r}

library(reshape2) #Loads the reshape2 package for transforming data format

#Calculates the number of flights that leave Newark each month by airline
num_flights_from_EWR <- flights_and_airlines %>%
  select(name, origin, month) %>%
  filter(origin == "EWR") %>%
  group_by(month, name) %>%
  summarise(Number.of.flights = length(name)) %>%
  dcast(name ~ month) %>%
  print()

```

* Using the **babynames** dataset:
    * Identify the ten most common male and female names in 2014. Make a plot of their frequency (`prop`) since 1880. (This may require two separate piped statements).
    
    * Make a single table of the 26th through 29th most common girls names in the year 1896, 1942, and 2016

```{r}

library(babynames) #Loads the babynames dataset

#Determines the ten most common male baby names of 2014
male_names14 <- babynames %>%
  filter(sex == "M", year == 2014) %>%
  arrange(rev(n)) %>%
  head(n = 10)
  
#Determines the ten most common female baby names of 2014
female_names14 <- babynames %>%
  filter(sex == "F", year == 2014) %>%
  arrange(rev(n)) %>%
  head(n = 10)

#Joins the male and female most popular names of 2014
pop_names14 <- full_join(male_names14, female_names14)
  
#Plots the frequency of the ten most common male and female baby names of 2014 since 1800
top10_babyname_freq_plot <- semi_join(babynames, pop_names14, by = c("name", "sex")) %>%
  ggplot(aes(year, prop, color = sex)) +
  facet_wrap(~name) +
  geom_point()

print(top10_babyname_freq_plot)

#Creates the table of common names as specified above
girl_names_26_29 <- babynames %>%
  filter(year == c(1896, 1942, 2016)) %>%
  group_by(year) %>%
  arrange(rev(n)) %>%
  slice(26:29) %>%
  select(year, name) %>%
  print()

```

*Note: There are no names for 2016 in the above table as the dataset end at 2014.*

* Write task that involves some of the functions on the Data Wrangling Cheat Sheet and execute it. You may either use your own data or data packages (e.g., the ones listed [here](https://blog.rstudio.org/2014/07/23/new-data-packages/)).

```{r}

#Creates a plot of the sex ratio of some unisex names over time
unisex_names_plot <- babynames %>%
  filter(name == c("Jessie", "Taylor", "Jackie")) %>%
  group_by(name, year) %>%
  mutate(total.per.year = sum(n)) %>%
  filter(sex == "M") %>%
  mutate(percent.male = (100 * n/total.per.year)) %>%
  filter(percent.male != 100) %>% #Filters out data for years where data is only present for one sex
  ggplot(aes(year, percent.male)) +
  facet_wrap(~name) +
  geom_point() +
  geom_smooth(method = lm)

print(unisex_names_plot)

```